\pkg{evtree}: Evolutionary Learning of Globally Optimal Classification and Regression Trees in \proglang{R}

### 2011-09-08 pkg/evtree/inst/doc/evtree.Rnw
cleanup of R code, figures, tables
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@225 255f4249-2263-4417-b1ea-8498779e27a7

%% Sweave/vignette information and metadata
%% need no \usepackage{Sweave}
\SweaveOpts{engine = R, eps = FALSE, keep.source = TRUE}
%\VignetteIndexEntry{Evolutionary Learning of Globally Optimal Classification and Regression Trees in R}
%\VignetteDepends{stats,evtree,partykit,mlbench,plotrix,lme4,multcomp,rpart,ctree,xtable}
%\VignetteKeywords{machine learning, classification trees, regression trees, evolutionary algorithms, R}
%\VignettePackage{evtree}

@@

\Abstract{
  Commonly used classification and regression tree methods like the CART
  algorithm are recursive partitioning methods that build the model in a forward
  stepwise search. Although this approach is known to be an efficient heuristic,
  the results of recursive tree methods are only locally optimal, as splits are
  chosen to maximize homogeneity at the next step only. An alternative way to
  search over the parameter space of trees is to use global optimization methods
  like evolutionary algorithms. This paper describes the \pkg{evtree} package,
  which implements an evolutionary algorithm for learning globally optimal
  classification and regression trees in \proglang{R}. CPU and memory-intensive
  tasks are fully computed in \proglang{C++} while the \pkg{partykit}
  \citep{hothorn2009partykit} package is leveraged to represent the resulting
  trees in \proglang{R}, providing unified infrastructure for summaries,
  visualizations, and predictions. \code{evtree} is compared to \code{rpart}
  \citep{therneau1997introduction}, the open-source CART implementation, and
  \code{ctree} \citep{hothorn2006urp} (conditional inference trees). On 20
  evaluated data-sets, including several benchmark problems from the UCI machine
  learning repository, \code{evtree} models offered at least similar and most of
  the time increased predictive accuracy.  
}

@@

\Address{
  Thomas Grubinger, Karl-Peter Pfeiffer\\
  Innsbruck Medical University\\
  Department for Medical Statistics, Informatics and Health Economics\\
  6020, Innsbruck, Austria\\
  E-mail: \email{thomas.grubinger@i-med.ac.at}\\
  URL: \url{http://www.i-med.ac.at/msig/mitarbeiter/grubinger}\\

@@

@@

The remainder of this paper is structured as follows. Section~\ref{got} describes the problem of learning globally optimal decision trees and contrasts it to the only locally optimal top-down heuristic that is utilized by recursive partitioning algorithms. In Section~\ref{desc} the \code{evtree} algorithm will be introduced. Section~\ref{impl} addresses implementation details and gives an overview of the implemented functions. In Section~\ref{comp} \code{evtree} is compared to the two commonly used algorithms \code{rpart} and \code{ctree}, from package \pkg{party}, which are the implementations of CART and conditional inference trees in \proglang{R}. The comparison comprises 14 benchmark data-sets, 3 real-world data-sets and 3 simulated data-sets. It will be shown that \code{evtree} models are often significantly more accurate compared to models from the two recursive partitioning methods. Finally, in Section~\ref{conc}, we will summarize and draw conclusions on the implementation and the performance of our algorithm.

@@

A very simple example that illustrates the limitation of recursive partitioning methods is depicted in Figure~\ref{fig:illustration}. The example only contains two independent variables and can be solved with only three splits that partition the input space into four regions. As expected the recursive partitioning methods \code{rpart} and \code{ctree} fail to find any split at all, as the impurity of the resulting subsets cannot be reduced by the first split. For methods that explore $\Theta$ in a more global fashion it is straightforward to find an optimal solution to this problem. One solution, which is the tree constructed by \code{evtree} is given by: 
%
<< echo = FALSE, split= TRUE>>=
library("evtree")
X1 <- c(rep(0.25, 4), rep(0.75, 4), rep(1.25 ,4),  rep(1.75, 4))
X2 <- c(rep(seq(0.25, 1.75 ,0.5),4))
Y <- array(1,16)
Y[X1 < 1 & X2 < 1]=2
Y[X1 > 1 & X2 > 1]=2
chess22 <- as.data.frame(cbind(X1, X2, Y))
chess22[,3] <- as.factor( chess22[,3] )
levels(chess22[,3]) = c("class: o","class: x")
print(evtree(Y~.,data=chess22, minbucket = 1, minsplit=2))
@

\setkeys{Gin}{width=0.4\textwidth}
\begin{figure}[tbh!]
\centering
<<fig = TRUE, echo = FALSE>>=
par(cex.axis=1.5, cex.lab=1.5)
plot(chess22[chess22[,3]=="class: o",1], chess22[chess22[,3]=="class: o",2], type="p",  pch= c("x"), col=colors()[320], xlab="X1", ylab="X2", xlim=c(0,2), ylim=c(0,2), cex=2)
points(chess22[chess22[,3]=="class: x",1], chess22[chess22[,3]=="class: x",2], pch= c("o"), cex=2)
par(cex.axis=1.5, cex.lab=1.5)
@
\caption{\label{fig:illustration} Class distribution of  the $(X_1, X_2)$-plane. The two classes are indicated by black $(o)$ and grey $(x)$.}
\end{figure}

All instances all are classified correctly. Each of the terminal nodes 3 and 7 contain four instances of the class $(x)$. Four instances of class $(o)$ are assigned to each of the terminal nodes 4 and 6.

@@

When compared with the described forward stepwise search, a less greedy approach is to calculate the effects of the choice of the split rule deeper down in the tree. In this way optimal trees can be found for simple problems. However, feature selection at a node of greedily induced trees has complexity  $O(PN)$. Through a search of the optimal split up to $d-1$ levels the complexity increases to $O(P^dN^d)$ \citep{papagelis2001bdt}. Therefore, for more difficult problems the search space at each node must be limited by the depth of the search. Additionally, the search space can be reduced, limiting the number of possible split points of the selected variables and/or other heuristics. These methods are known as \textit{look-ahead algorithms} \citep{esmeir2007ald}. Another class of algorithms is given by stochastic optimization methods. Given an initial tree, these methods seek improved solutions through stochastic changes to the tree structure. Besides evolutionary algorithms \citep{koza1991concept}, \textit{Bayesian CART} \citep{denison1998bca} and \textit{simulated annealing} \citep{sutton1992ict} were used successfully to solve difficult classification and regression tree problems. \cite{koza1991concept} first formulated the concept of using evolutionary algorithms as a stochastic optimization method to build classification and regression trees. \cite{papagelis2001bdt} presented a classification tree algorithm and provided results on several data-sets from the UCI machine learning repository \citep{Frank+Asuncion:2010}. A method for the construction of regression trees via evolutionary algorithms was introduced by \cite{fan2005rta}.  \cite{cantu2003inducing} used an evolutionary algorithm to induce oblique classification trees. \cite{gray2008cta} introduced a classification tree algorithm that allows univariate splits, but also splits defined by linear combinations of two or three variables.

@@

\begin{table}[tbh!]
\begin{center}
\begin{tabular}{l}
\toprule
1. initialize the population \\
2. evaluate each individual  \\
3. while(termination condition is not satisfied) do  \\
\hspace{8mm} a. select parents  \\
\hspace{8mm} b. alter selected individuals via variation operators \\
\hspace{8mm} c. evaluate new solutions  \\
\hspace{8mm} d. select survivors for the next generation  \\
\hspace{3.8mm} od  \\
\bottomrule
\end{tabular}
\end{center}
\caption{Pseudocode of the general schema of an evolutionary algorithm.} 
\label{pc:genericEA}
\end{table}

@@

Other, existing and commonly used choices of evaluation functions include the \textit{Bayesian Information Criteria (BIC)} \citep{gray2008cta} and \textit{Minimum Description Length (MDL)} \citep{quinlan1989inferring}. For both evaluation functions deviance is used for accuracy estimation. Deviance is usually preferred over the misclassification rate in recursive partitioning methods, as it is more sensitive to changes in the node probabilities \citep[pp.~308--310]{hastie}. However this is not necessarily an advantage for global tree building methods like evolutionary algorithms.

@@

For regression trees, accuracy is usually measured by the mean squared error $MSE(Y, f(X, \theta))= \sum_{n=1}^N (Y_n - f(X_{\cdot n}, \theta)^2)$. Using $N \cdot \ln MSE(Y, f(X, \theta))$ as a loss function and $\alpha \cdot 4 \cdot (M+1) \cdot \ln N$ as the complexity part, the general formulation of the optimization problem in Equation~\ref{generalcostfunction} can be rewritten as:

\begin{equation}
	 \hat{\theta} = \operatorname*{\arg \,\min}_{\theta \in \Theta} \; N \cdot \ln MSE(Y, f(X, \theta))+\alpha \cdot 4 \cdot (M+1) \cdot \ln N.
\label{eq:regression}
\end{equation}

@@

\code{evtree} provides an efficient implementation of an evolutionary algorithm that builds classification trees in \proglang{R}. CPU- and memory- intensive tasks are fully computed in \proglang{C++}, while the user interfaces and plot functions are written in \proglang{R}. The \code{.C()} interface \citep{chambers2008software} was used to pass arguments between the two languages. \code{evtree} depends on the \pkg{partykit} package  \citep{hothorn2009partykit}, which provides an infrastructure for representing, summarizing, and visualizing tree-structured models. The supplied functionality includes standard \code{print()}, \code{summary()}, and \code{plot()} functions to display trees and a \code{predict()} function to predict the class labels of new data. The arguments of  \code{evtree()} are
%
\begin{Code}
  evtree(formula, data = list(), weights = NULL, subset = NULL,
    control = evtree.control(...), ...)
\end{Code}
%
where \code{formula} and \code{data} are a standard \proglang{R} formula of type \code{y ~ x1 + x2} with the respective data. \code{weights} is an optional vector specifying weights to be used in the fitting process. The option vector subset specifies a subset of observations to be used in the fitting process. \code{control} comprises a list of control parameters
%
\begin{Code}
  evtree.control(minbucket = 7L, minsplit = 20L, maxdepth = 9L, 
    niterations = 10000L, ntrees = 100L, alpha = 1,
    operatorprob = list(pmutatemajor = 0.2, pmutateminor = 0.2,
      pcrossover = 0.2, psplit = 0.2, pprune = 0.2),
    seed = NULL, ...)
\end{Code}
%
where the parameters \code{minbucket}, \code{minsplit}, and \code{maxdepth} constrain the solution to a minimum number of observations in each terminal node, a minimum number of observation in each internal node and a maximum tree depth. Note that the memory requirements increase by the square of the maximum tree depth. Parameter \code{alpha} regulates the complexity part of the evaluation function and refers to $\alpha$ in Equation~\ref{eq:classification} for classification trees and Equation~\ref{eq:regression} for regression trees. \code{niterations} and \code{ntrees} specify the maximum number of iterations and the number of trees in the population. With the argument \code{operatorprob}, user-specified probabilities for the variation operators can be defined. Argument \code{seed} is an optional seed for the random number generator, which allows models to be reproduced by re-using the same seed. If not specified, the random number generator is initialized by the system time. In this case, iterative calls of \code{evtree} will possibly result in different models.
%
<<>>=
library("evtree")
data("iris")
model <- evtree(Species ~ ., iris, control = evtree.control(seed = 1000))
@
%
The fitted model can be examined using the \code{print()} and \code{plot()} functions. 
%
<<>>=
print(model)
@

\setkeys{Gin}{width=1\textwidth}
\begin{figure}[t!]
\centering
<<echo= FALSE>>=
par(cex.axis=1.5, cex.lab=1.5)
<<fig=TRUE, echo=TRUE, width = 13.5, height= 8.5>>=
plot(model)
@
\caption{\label{fig:example} Example of a tree plotted by the \code{plot()} function.}
\end{figure}

The resulting plot is displayed in Figure~\ref{fig:example}. \code{predict()} can be used for the prediction of the response.
%
<<>>=
predict(model, newdata = iris[c(10, 35, 60, 85, 110, 135), -5])
@

@@

In the first part of the analysis (Section~\ref{comp1}) the algorithms are compared on 14 benchmark data-sets that are publicly available and $3$ real-world data-sets from the Austrian \textit{Diagnosis Related Group (DRG)} system \citep{LKF2009}. The analysis is based on the evaluation of 250 bootstrap samples for each of the 20 data-sets. The misclassification rate on the \textit{out-of-bag} \citep{hothorn2005design} samples is used as a measure of predictive accuracy. Furthermore, the complexity is estimated by the number of terminal nodes. Missing values are currently not supported by \code{evtree}. The only data-set with missing values is the \textit{Breast Cancer Database} with a total of 16 missing values, which were removed before analysis. Dunnett's correction from \proglang{R} package \pkg{multcomp} \citep{multcomp} was used to calculate simultaneous $95\%$ confidence intervals for the individual data-sets .

@@

In Table~\ref{tab:dataDescription} the benchmark and real-world data-sets from the Austrian DRG-system are described. In the Austrian DRG-system, resources are allocated to hospitals by simple rules regarding diagnosis, selected individual services, and patient characteristics. Within this system regression tree analysis is performed to classify hospital patients into clinically meaningful and comprehensible groups that consume similar hospital resources. The data-sets MEL0101 (intervention on the skull), HDG0202 (malignant neoplasms), and HDG0502 (acute affections of the respiratory tract and middle atelectasis) contain patient groups with the same main diagnosis or procedure. The task of the regression tree analysis is to further divide these groups with the intention of finding groups with similar resource consumption. Possible split-variables are the patients' diagnoses, procedures, age, and gender. The dependent variable is the patients' length of stay, which has a good relationship with the total costs and is available \citep{theurl2007ihf}. Permission to use the data was granted by the \citet{BMG2009}. A more detailed description of the application can be found in \citet{grubinger2010regression}.

\begin{table}
\begin{center}
\begin{tabular}{lrrrrrr}
\toprule
\textsc{Data-set} & \textsc{Instances} &  \multicolumn{5}{l}{\textsc{Attributes}}   \\ 
                  &                    &  binary              & nominal          &  ordered      &  metric     & classes \\ 
\midrule
Glass Identification$^\#$  &   214   &  - &  - &  - & 9 &  6      	 \\
Statlog Heart*         &   270   &  3 &   3 & 1 & 6 &   2     	 \\
Ionosphere$^\#$ &   351   &  2 &  - &  - & 32 & 2  	             \\
Musk$^+$  &  476  & - &  - &  - & 166 & 2                     	 \\
Breast Cancer Database$^\#$ &   685   &  - &  4 & 5 &  - &  2  	 \\
Pima Indians Diabetes$^\#$ &   768   &  - &   - & - & 8 &  2      	 \\
Vowel$^\#$  &  990  & - &  1 &  - & 9 & 11     	                     \\
Statlog German Credit* &  1000   &  2 &  10 & 1 & 7 &  2      	 \\
Contraceptive Method* &   1437   & 3 &  - &  4 &  2 &  3  	     \\
DNA$^\#$  &  3186  & 180 &  - &  - & - & 3   	                       \\
Spam$^+$  &  4601  & - &  - &  - & 57 & 2                      	 \\
MAGIC Gamma Telescope*  &  19020  & - &  - &  - & 10 & 2     	   \\
Servo$^\#$ & 167 &  - & 4 & - &  -  &  -											   \\
Boston Housing$^\#$ & 506 & 1 & - & - &  12  & -                 \\
MEL0101$^\diamond$ &   875 &  1 & 4 & 1 &  108  & -              \\
HDG0202$^\diamond$ &  3933 &  1 & 7 & 1 &   46  & -              \\
HDG0502$^\diamond$ & 8251 &  1 & 7 & 1 &   91  & -               \\
\bottomrule
\end{tabular}
\end{center}
\caption{Description of the evaluated benchmark data-sets. The data-sets marked with a $(^*)$ are from the UCI machine learning repository \citep{Frank+Asuncion:2010}. Data-sets marked with a $(^\#)$ and $(^{+})$ are from the \proglang{R} packages \pkg{mlbench} \citep{mlbench} and \pkg{kernlab} \citep{kernlab}. The $3$ real-world data-sets from the Austrian DRG system are marked with a $(^\diamond)$. } 
\label{tab:dataDescription}
\end{table}

The relative performance of \code{evtree} and \code{rpart} is summarized in Figure~\ref{fig:crpart}. Performance differences are displayed relative to \code{evtree}'s performance. For example, on the \textit{Glass} data-set, the average misclassification rate of \code{rpart} is $2.7\%$ higher than the misclassification rate of \code{evtree}. It can be observed that on $12$ out of $17$ data-sets \code{evtree} significantly outperforms \code{rpart} in terms of predictive accuracy. Only on the \textit{Contraceptive Method} data-set \code{evtree} performs slightly worse.  In terms of complexity, \code{evtree} models are significantly more complex on $10$ and less complex on $7$ data-sets.

Figure~\ref{fig:cptree} summarizes the relative performance of \code{evtree} and \code{ctree}. For 15 out of 17 data-sets \code{evtree} shows a better predictive performance. The algorithms' performances  is significantly worse on the \textit{MEL0101} data-set, where the average misclassifiation rate of \code{ctree} is $5.6\%$ lower. However, on this data-set, \code{ctree} models are on average $86.5\%$ larger than \code{evtree} models. The relative complexity of \code{evtree} models is significantly smaller for 15 and larger for 1 data-set.

%% vorberechnungen
<<label=figciplots, echo = FALSE>>=
rm(list = ls())
library("plotrix")
library("lme4")
library("multcomp")
load(file="./results/breastcancer.RData")
load(file="./results/glass.RData")
load(file="./results/credit.RData")
load(file="./results/heart.RData")
load(file="./results/pima.RData")
load(file="./results/contraceptive.RData")
load(file="./results/dna.RData")
load(file="./results/musk.RData")
load(file="./results/spam.RData")
load(file="./results/vowel.RData")
load(file="./results/ionosphere.RData")
load(file="./results/magicgamma.RData")
load(file="./results/hdg0202.RData")
load(file="./results/hdg0502.RData")
load(file="./results/mel0101.RData")
load(file="./results/bostonhousing.RData")
load(file="./results/servo.RData")

>>>>

%% Sweave/vignette information and metadata
%% need no \usepackage{Sweave}
\SweaveOpts{engine = R, eps = FALSE, keep.source = TRUE}
%\VignetteIndexEntry{Evolutionary Learning of Globally Optimal Classification and Regression Trees in R}
%\VignetteDepends{stats,evtree,partykit,mlbench,multcomp,rpart,party,xtable}
%\VignetteKeywords{machine learning, classification trees, regression trees, evolutionary algorithms, R}
%\VignettePackage{evtree}

@@

\Abstract{
  Commonly used classification and regression tree methods like the CART
  algorithm are recursive partitioning methods that build the model in a forward
  stepwise search. Although this approach is known to be an efficient heuristic,
  the results of recursive tree methods are only locally optimal, as splits are
  chosen to maximize homogeneity at the next step only. An alternative way to
  search over the parameter space of trees is to use global optimization methods
  like evolutionary algorithms. This paper describes the \pkg{evtree} package,
  which implements an evolutionary algorithm for learning globally optimal
  classification and regression trees in \proglang{R}. CPU and memory-intensive
  tasks are fully computed in \proglang{C++} while the \pkg{partykit}
  \citep{hothorn2009partykit} package is leveraged to represent the resulting
  trees in \proglang{R}, providing unified infrastructure for summaries,
  visualizations, and predictions. \code{evtree} is compared to \code{rpart}
  \citep{therneau1997introduction}, the open-source CART implementation, and
  \code{ctree} \citep{hothorn2006urp} (conditional inference trees). On 20
  evaluated datasets, including several benchmark problems from the UCI machine
  learning repository, \code{evtree} models offered at least similar and most of
  the time increased predictive accuracy.  
}

@@

\Address{
  Thomas Grubinger, Karl-Peter Pfeiffer\\
  Innsbruck Medical University\\
  Department for Medical Statistics, Informatics and Health Economics\\
  6020, Innsbruck, Austria\\
  E-mail: \email{Thomas.Grubinger@i-med.ac.at}, \email{Karl-Peter.Pfeiffer@i-med.ac.at}\\
  URL: \url{http://www.i-med.ac.at/msig/mitarbeiter/grubinger/}\\
  \phantom{URL: }\url{http://www.i-med.ac.at/msig/mitarbeiter/pfeiffer/}\\

@@

<<setup, echo=FALSE>>=
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
library("evtree")
@

@@

The remainder of this paper is structured as follows. Section~\ref{got} describes the problem of learning globally optimal decision trees and contrasts it
to the only locally optimal top-down heuristic that is utilized by recursive partitioning algorithms. In Section~\ref{desc} the \code{evtree} algorithm will
be introduced. Section~\ref{impl} addresses implementation details and gives an overview of the implemented functions. In Section~\ref{comp} \code{evtree}
is compared to the two commonly used algorithms \code{rpart} and \code{ctree}, from package \pkg{party}, which are the implementations of CART and
conditional inference trees in \proglang{R}. The comparison comprises 14~benchmark datasets, 3~real-world datasets and 3~simulated datasets. It will be shown that \code{evtree} models are often significantly more accurate compared to models from the two recursive partitioning methods. Finally, in Section~\ref{conc}, we will summarize and draw conclusions on the implementation and the performance of our algorithm.

@@

A very simple example that illustrates the limitation of recursive partitioning methods is depicted in Figure~\ref{fig:illustration}. The example only contains two independent variables and can be solved with only three splits that partition the input space into four regions. As expected the recursive partitioning methods \code{rpart} and \code{ctree} fail to find any split at all, as the impurity of the resulting subsets cannot be reduced by the first split. For methods that explore $\Theta$ in a more global fashion it is straightforward to find an optimal solution to this problem. One solution, which is the tree constructed by \code{evtree} is given by: 
%
<<chess22, echo=FALSE>>=
X1 <- rep(seq(0.25, 1.75, 0.5), each = 4)
X2 <- rep(seq(0.25, 1.75, 0.5), 4)
Y <- rep(1, 16)
Y[(X1 < 1 & X2 < 1) | (X1 > 1 & X2 > 1)] <- 2
Y <- factor(Y, labels = c("O", "X"))
chess22 <- data.frame(Y, X1, X2)
print(evtree(Y ~ ., data = chess22, minbucket = 1, minsplit = 2))
@

\setkeys{Gin}{width=0.5\textwidth}
\begin{figure}[t!]
\centering
<<chess22-plot, fig=TRUE, echo=FALSE, height=4, width=4>>=
plot(X2 ~ X1, data = chess22, xlim = c(0, 2), ylim = c(0, 2), pch = c(1, 4)[Y], col = c("black", "slategray")[Y])
@
\caption{\label{fig:illustration} Class distribution of  the $(X_1, X_2)$-plane. The two classes are indicated by black circles and gray crosses.}
\end{figure}

All instances all are classified correctly. Each of the terminal nodes~3 and~7 contain four instances of the class \code{X}. Four instances of class
\code{O} are assigned to each of the terminal nodes~4 and~6.

@@

When compared with the described forward stepwise search, a less greedy approach is to calculate the effects of the choice of the split rule deeper down in the tree. In this way optimal trees can be found for simple problems. However, feature selection at a node of greedily induced trees has complexity  $O(PN)$. Through a search of the optimal split up to $d-1$ levels the complexity increases to $O(P^dN^d)$ \citep{papagelis2001bdt}. Therefore, for more difficult problems the search space at each node must be limited by the depth of the search. Additionally, the search space can be reduced, limiting the number of possible split points of the selected variables and/or other heuristics. These methods are known as \textit{look-ahead algorithms} \citep{esmeir2007ald}. Another class of algorithms is given by stochastic optimization methods. Given an initial tree, these methods seek improved solutions through stochastic changes to the tree structure. Besides evolutionary algorithms \citep{koza1991concept}, \textit{Bayesian CART} \citep{denison1998bca} and \textit{simulated annealing} \citep{sutton1992ict} were used successfully to solve difficult classification and regression tree problems. \cite{koza1991concept} first formulated the concept of using evolutionary algorithms as a stochastic optimization method to build classification and regression trees. \cite{papagelis2001bdt} presented a classification tree algorithm and provided results on several datasets from the UCI machine learning repository \citep{Frank+Asuncion:2010}. A method for the construction of regression trees via evolutionary algorithms was introduced by \cite{fan2005rta}.  \cite{cantu2003inducing} used an evolutionary algorithm to induce oblique classification trees. \cite{gray2008cta} introduced a classification tree algorithm that allows univariate splits, but also splits defined by linear combinations of two or three variables.

@@

\begin{table}[t!]
\centering
\hrulefill
\begin{enumerate}
  \item Initialize the population.
  \item Evaluate each individual.
  \item While(termination condition is not satisfied) do:
  \begin{itemize}
    \item[a.] Select parents.
    \item[b.] Alter selected individuals via variation operators.
    \item[c.] Evaluate new solutions.
    \item[d.] Select survivors for the next generation.
  \end{itemize}
\end{enumerate}
\hrulefill
\caption{Pseudocode of the general schema of an evolutionary algorithm.} 
\label{pc:genericEA}
\end{table}

@@

Other, existing and commonly used choices of evaluation functions include the \textit{Bayesian information criterion} \citep[BIC, as in][]{gray2008cta} and
\textit{minimum description length} \citep[MDL, as in][]{quinlan1989inferring}. For both evaluation functions deviance is used for accuracy estimation. Deviance is usually preferred over the misclassification rate in recursive partitioning methods, as it is more sensitive to changes in the node probabilities \citep[pp.~308--310]{hastie}. However this is not necessarily an advantage for global tree building methods like evolutionary algorithms.

@@

For regression trees, accuracy is usually measured by the mean squared error $\mathit{MSE}(Y, f(X, \theta))= \sum_{n=1}^N (Y_n - f(X_{\cdot n}, \theta)^2)$. Using $N \cdot \ln \mathit{MSE}(Y, f(X, \theta))$ as a loss function and $\alpha \cdot 4 \cdot (M+1) \cdot \ln N$ as the complexity part, the general formulation of the optimization problem in Equation~\ref{generalcostfunction} can be rewritten as:

\begin{equation}
	 \hat{\theta} = \operatorname*{\arg \,\min}_{\theta \in \Theta} \; N \cdot \ln \mathit{MSE}(Y, f(X, \theta))+\alpha \cdot 4 \cdot (M+1) \cdot \ln N.
\label{eq:regression}
\end{equation}

@@

\code{evtree} provides an efficient implementation of an evolutionary algorithm that builds classification trees in \proglang{R}. CPU- and memory- intensive tasks are fully computed in \proglang{C++}, while the user interfaces and plot functions are written in \proglang{R}. The \code{.C()} interface \citep{chambers2008software} was used to pass arguments between the two languages. \code{evtree} depends on the \pkg{partykit} package  \citep{hothorn2009partykit}, which provides an infrastructure for representing, summarizing, and visualizing tree-structured models. The supplied functionality includes standard \code{print()}, \code{summary()}, and \code{plot()} functions to display trees and a \code{predict()} function to predict the class labels of new data. The arguments of  \code{evtree()} are
%
\begin{Code}
  evtree(formula, data = list(), weights = NULL, subset = NULL,
    control = evtree.control(...), ...)
\end{Code}
%
where \code{formula} and \code{data} are a standard \proglang{R} formula of type \code{y ~ x1 + x2} with the respective data. \code{weights} is an optional vector specifying weights to be used in the fitting process. The option vector subset specifies a subset of observations to be used in the fitting process. \code{control} comprises a list of control parameters
%
\begin{Code}
  evtree.control(minbucket = 7L, minsplit = 20L, maxdepth = 9L, 
    niterations = 10000L, ntrees = 100L, alpha = 1,
    operatorprob = list(pmutatemajor = 0.2, pmutateminor = 0.2,
      pcrossover = 0.2, psplit = 0.2, pprune = 0.2),
    seed = NULL, ...)
\end{Code}
%
where the parameters \code{minbucket}, \code{minsplit}, and \code{maxdepth} constrain the solution to a minimum number of observations in each terminal node, a minimum number of observation in each internal node and a maximum tree depth. Note that the memory requirements increase by the square of the maximum tree depth. Parameter \code{alpha} regulates the complexity part of the evaluation function and refers to $\alpha$ in Equation~\ref{eq:classification} for classification trees and Equation~\ref{eq:regression} for regression trees. \code{niterations} and \code{ntrees} specify the maximum number of iterations and the number of trees in the population. With the argument \code{operatorprob}, user-specified probabilities for the variation operators can be defined. Argument \code{seed} is an optional seed for the random number generator, which allows models to be reproduced by re-using the same seed. If not specified, the random number generator is initialized by the system time. In this case, iterative calls of \code{evtree} will possibly result in different models.
%
<<iris-tree>>=
library("evtree")
data("iris", package = "datasets")
ev <- evtree(Species ~ ., iris, control = evtree.control(seed = 1000))
@
%
The fitted model can be examined using the \code{print()} and \code{plot()} functions. 
%
<<iris-print>>=
ev
plot(ev)
@

\setkeys{Gin}{width=1\textwidth}
\begin{figure}[t!]
\centering
<<iris-plot, fig=TRUE, echo=FALSE, width=13.2, height=8>>=
plot(ev)
@
\caption{\label{fig:example} Example of a tree plotted by the \code{plot()} function.}
\end{figure}

The resulting plot is displayed in Figure~\ref{fig:example}. \code{predict()} can be used for the prediction of the response.
%
<<iris-predict>>=
predict(ev, newdata = iris[c(10, 35, 60, 85, 110, 135), -5])
@

@@

In the first part of the analysis (Section~\ref{comp1}) the algorithms are compared on 14 benchmark datasets that are publicly available and $3$ real-world datasets from the Austrian \textit{Diagnosis Related Group (DRG)} system \citep{LKF2009}. The analysis is based on the evaluation of 250 bootstrap samples for each of the 20 datasets. The misclassification rate on the \textit{out-of-bag} \citep{hothorn2005design} samples is used as a measure of predictive accuracy. Furthermore, the complexity is estimated by the number of terminal nodes. Missing values are currently not supported by \code{evtree}. The only dataset with missing values is the \textit{Breast Cancer Database} with a total of 16 missing values, which were removed before analysis. Dunnett's correction from \proglang{R} package \pkg{multcomp} \citep{multcomp} was used to calculate simultaneous $95\%$ confidence intervals for the individual datasets .

@@

In Table~\ref{tab:dataDescription} the benchmark and real-world datasets from the Austrian DRG-system are described. In the Austrian DRG-system, resources are allocated to hospitals by simple rules regarding diagnosis, selected individual services, and patient characteristics. Within this system regression tree analysis is performed to classify hospital patients into clinically meaningful and comprehensible groups that consume similar hospital resources. The datasets MEL0101 (intervention on the skull), HDG0202 (malignant neoplasms), and HDG0502 (acute affections of the respiratory tract and middle atelectasis) contain patient groups with the same main diagnosis or procedure. The task of the regression tree analysis is to further divide these groups with the intention of finding groups with similar resource consumption. Possible split-variables are the patients' diagnoses, procedures, age, and gender. The dependent variable is the patients' length of stay, which has a good relationship with the total costs and is available \citep{theurl2007ihf}. Permission to use the data was granted by the \citet{BMG2009}. A more detailed description of the application can be found in \citet{grubinger2010regression}.

\begin{table}[t!]
\centering
\begin{tabular}{lrrrrrr}
\toprule
{Dataset} & {Instances} &  \multicolumn{5}{l}{{Attributes}}   \\ 
           &             &  Binary              & Nominal          &  Ordered      &  Metric     & Classes \\ 
\midrule
Glass identification$^\#$  &   214   &  - &  - &  - & 9 &  6      	 \\
Statlog heart*         &   270   &  3 &   3 & 1 & 6 &   2     	 \\
Ionosphere$^\#$ &   351   &  2 &  - &  - & 32 & 2  	             \\
Musk$^+$  &  476  & - &  - &  - & 166 & 2                     	 \\
Breast cancer database$^\#$ &   685   &  - &  4 & 5 &  - &  2  	 \\
Pima Indians diabetes$^\#$ &   768   &  - &   - & - & 8 &  2      	 \\
Vowel$^\#$  &  990  & - &  1 &  - & 9 & 11     	                     \\
Statlog German credit* &  1000   &  2 &  10 & 1 & 7 &  2      	 \\
Contraceptive method* &   1437   & 3 &  - &  4 &  2 &  3  	     \\
DNA$^\#$  &  3186  & 180 &  - &  - & - & 3   	                       \\
Spam$^+$  &  4601  & - &  - &  - & 57 & 2                      	 \\
MAGIC gamma telescope*  &  19020  & - &  - &  - & 10 & 2     	   \\
Servo$^\#$ & 167 &  - & 4 & - &  -  &  -											   \\
Boston housing$^\#$ & 506 & 1 & - & - &  12  & -                 \\
MEL0101$^\diamond$ &   875 &  1 & 4 & 1 &  108  & -              \\
HDG0202$^\diamond$ &  3933 &  1 & 7 & 1 &   46  & -              \\
HDG0502$^\diamond$ & 8251 &  1 & 7 & 1 &   91  & -               \\
\bottomrule
\end{tabular}
\caption{\label{tab:dataDescription} Description of the evaluated benchmark datasets. The datasets marked with
  $*$ originate from the UCI machine learning repository \citep{Frank+Asuncion:2010} and are made available in the \pkg{evtree} package.
  Datasets marked with $\#$ and $+$ are from the \proglang{R} packages \pkg{mlbench} \citep{mlbench} and
  \pkg{kernlab} \citep{kernlab}, respectively. The three real-world datasets from the Austrian DRG system are marked with $\diamond$.} 
\end{table}

The relative performance of \code{evtree} and \code{rpart} is summarized in Figure~\ref{fig:benchmark} (upper panels). Performance differences are displayed relative to \code{evtree}'s performance. For example, on the \textit{Glass} dataset, the average misclassification rate of \code{rpart} is $2.7\%$ higher than the misclassification rate of \code{evtree}. It can be observed that on $12$ out of $17$ datasets \code{evtree} significantly outperforms \code{rpart} in terms of predictive accuracy. Only on the \textit{Contraceptive Method} dataset \code{evtree} performs slightly worse.  In terms of complexity, \code{evtree} models are significantly more complex on $10$ and less complex on $7$ datasets.

Figure~\ref{fig:benchmark} (lower panels) summarizes the relative performance of \code{evtree} and \code{ctree}. For 15 out of 17 datasets \code{evtree} shows a better predictive performance. The algorithms' performances  is significantly worse on the \textit{MEL0101} dataset, where the average misclassifiation rate of \code{ctree} is $5.6\%$ lower. However, on this dataset, \code{ctree} models are on average $86.5\%$ larger than \code{evtree} models. The relative complexity of \code{evtree} models is significantly smaller for 15 and larger for 1 dataset.

<<benchmark-results, echo=FALSE>>=
## load results
for(i in Sys.glob("results/*.RData")) load(i)

## preprocess for reference evtree
preprocess <- function(d, dname = "datasetname", isclassification = TRUE){
    if(isclassification) d[, 1:3] <- 1 - d[ ,1:3]
    d <- as.data.frame(d)
    colnames(d) <- c("evtree", "rpart", "ctree","evtree", "rpart", "ctree")
    for(i in 3:1) d[, i] <- d[, i] / d[, 1] * 100
    for(i in 6:4) d[, i] <- d[, i] / d[, 4] * 100
    x <- d[, 1:3]
    y <- d[, 4:6]
    rval <- reshape(x, idvar="samp", times=names(x), timevar = "alg",varying= list(names(x)), direction="long")
    names(rval)[2] <- "accuracy"
    rval$complexity <- reshape(y, idvar="samp", times=names(y), timevar = "alg",varying= list(names(y)), direction="long")[,2]
    rval$alg <- factor(rval$alg, levels = c("evtree", "ctree", "rpart"))
    rval$ds <- dname
    rval
}
